---
output: github_document
---

# OMIP 030

```{r global-options, include=FALSE}
library(knitr)
opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r libraries}
library(magrittr)
library(tidyverse)
library(fs)
library(janitor)
library(FlowRepositoryR)
library(flowCore)
library(openCyto)
library(ggcyto)
library(Rtsne)
library(reshape2)

map <- purrr::map
add <- flowCore::add
filter <- dplyr::filter
```

```{r theme, include=FALSE}
# Set common ggplot theme
flow_theme <- theme_bw() + theme(
  panel.grid.minor.x=element_blank(),
  panel.grid.minor.y=element_blank(),
  panel.grid.major.x=element_blank(),
  panel.grid.major.y=element_blank()
)
```

Load a flow workspace, which will contain all gaiting and cell expression information:

```{r load_workspace}
ws <- openWorkspace('~/analysis/data/omip/omip-030/clean/OMIP_PBMC_clean.wsp')
gs <- parseWorkspace(ws, name='TCells', 
                     path='/home/eczech/analysis/data/omip/omip-030/clean', isNcdf=FALSE)
# Extract single GatingHierarchy from GatingSet
gh <- gs[[1]]
```

```{r summarize_flow_frame, results='markup'}
print(getData(gh))
```

Plot the gating workflow starting from where T Cells are first identified:

```{r show_workflow, fig.width=12}
plot(gs, 'T Cells', width=6, height=1, fontsize=18, shape='rectangle')
```


Identify terminal nodes in the workflow which were, by convention, named with either ```CD4+``` or ```CD8+``` prefixed to their population names:

```{r node_groups}
all_nodes <- getNodes(gs, path=1)
cell_pops <- all_nodes %>% keep(str_detect(., '^CD[4|8]\\+ .*'))

not.is.na <- function(x) !is.na(x)
cell_pop_to_class <- function(x) case_when(
  str_detect(x, '^CD4\\+') ~ 'CD4+',
  str_detect(x, '^CD8\\+') ~ 'CD8+',
  TRUE ~ NA_character_
)
cell_pop_to_mem_state <- function(x) case_when(
  str_detect(x, '^CD[4|8]\\+ CM.*') ~ 'CM',
  str_detect(x, '^CD[4|8]\\+ EM.*') ~ 'EM',
  TRUE ~ NA_character_
)
  
cell_mem_types <- cell_pops[cell_pops %>% cell_pop_to_mem_state %>% not.is.na]
cell_pheno_types <- cell_pops %>% discard(~. %in% cell_mem_types)
```

Show the gating for terminal nodes, which correspond to specific cell types:

```{r leaf_gates, fig.width=12, fig.height=12}
autoplot(gs[[1]], cell_pops, bins=100, strip.text='gate', axis_inverse_trans=FALSE) + 
  ggcyto_par_set(facet=facet_wrap(~name, scales='free'), limits = list(x=c(0, 225), y=c(0, 225)))
```

Plot the distribution of different cell types noting that the memory phenotype (CM vs EM) exhibited by CD4+ and CD8+ cells is considered independent of the other T-helper, Regulatory, Effector and Naive phenotypes.  For example, this means that a Th17 cell can also exhibit CM or EM cell markers (or neither).  Before separating these designations, this distribution will show how frequent each phenotype is regardless of co-occurrence:

```{r cell_type_distribution, fig.width=12}
cell_count <- getTotal(gh, 'T Cells')

cell_pop <- getPopStats(gs) %>% 
  filter(Population %in% cell_pops) %>% 
  select(Population, Count) %>% 
  arrange(Count)

cell_pop %>%
  mutate(Percent=Count/cell_count, Class=cell_pop_to_class(Population)) %>% 
  arrange(desc(Percent)) %>% 
  mutate(Population=fct_reorder(Population, Percent)) %>%
  ggplot(aes(x=Population, y=Percent, fill=Class)) +
  geom_bar(stat='identity') +
  scale_fill_brewer(palette='Set1') +
  scale_y_continuous(labels = scales::percent) +
  ggtitle('T Cell Type Distribution') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

Now assign a primary phenotype and a memory phenotype to each cell and show how frequent either occur together:

```{r phenotype_distribution, results='asis'}
extract_node_assignment <- function(nodes) {
  nodes %>% map(~getIndices(gh, .)) %>% bind_cols %>% 
    set_names(nodes) %>% apply(., 1, function(x){
      if (sum(x) == 1) nodes[x] 
      else NA_character_
    })
}

df_cell_types <- list(pheno_type=cell_pheno_types, mem_type=cell_mem_types) %>% 
  map(extract_node_assignment) %>% as_tibble %>%
  mutate(class_type=cell_pop_to_class(pheno_type)) %>%
  mutate(pheno_type=case_when(
    pheno_type %in% c('CD4+ Th2g1', 'CD4+ Th2g2') ~ 'CD4+ Th2', 
    TRUE ~ pheno_type
  ))

df_cell_types %>% filter(!is.na(pheno_type)) %>% 
  mutate(mem_type=case_when(is.na(mem_type) ~ 'None', TRUE ~ mem_type)) %>% 
  tabyl(pheno_type, mem_type) %>% 
  adorn_totals('row') %>% adorn_totals('col') %>%
  adorn_percentages(denominator='all') %>% 
  adorn_pct_formatting(digits=1) %>%
  knitr::kable()
```

Given the above information, we're ready to extract the raw marker intensity data and attach enough cell metadata to it to do something useful:

```{r extract}
# Create data frame with measurments and phenotype information
fr <- getData(gh)
name_map <- parameters(fr)@data %>% filter(str_detect(name, 'Comp'))
name_map <- set_names(name_map$name, name_map$desc)
df <- fr %>% exprs %>% as_tibble %>% 
  select(starts_with('Comp')) %>% 
  rename(!!name_map) %>%
  cbind(df_cell_types) %>%
  filter(!is.na(pheno_type))
```


Now that the data is in a convenient form, one way to analyze this is to look at how well any one marker works to separate a primary phenotype from all others.  Below, a Kolmogrov Smirnov statistic signed by median difference is used where for any one marker, all cells of a specific primary phenotype are compared to all cells of a different phenotype to give an overall sense of magnitude and direction of the separation:

```{r ks, fig.width=12, fig.height=8}
dfg <- df %>% group_by(pheno_type) %>% do({
  d <- .
  dp <- df %>% filter(pheno_type != d$pheno_type[1]) %>% select_if(is.numeric)
  dt <- d %>% select_if(is.numeric)
  assertthat::are_equal(colnames(dp), colnames(dt))
  stats <- lapply(colnames(dp), function(col) {
    x <- pull(dt, col)
    y <- pull(dp, col)
    sign(median(x) - median(y)) * ks.test(x, y)$statistic[[1]]
  })
  names(stats) <- colnames(dp)
  data.frame(stats)
}) %>% ungroup

hc <- dfg %>% select_if(is.numeric) %>% as.matrix %>% dist %>% hclust
dfg %>% melt(id.vars='pheno_type') %>% 
  mutate(pheno_type=factor(pheno_type, levels=dfg$pheno_type[hc$order])) %>% 
  ggplot(aes(x=pheno_type, y=variable, fill=value)) +
  geom_tile() + 
  scale_fill_gradient2(
    low='darkred', mid='white', high='darkblue', 
    guide=guide_colorbar(title='KS Statistic')) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab('Primary Phenotype') + ylab('Marker') + 
  ggtitle('Expression Heatmap')
```

TSNE decomposition shows how much more readily some phenotypes separate from others:

```{r tsne, fig.width=12, fig.height=8}
set.seed(1)
df_samp <- df %>% sample_n(15000)
tsne <- df_samp %>% select_if(is.numeric) %>% as.matrix %>% Rtsne

df_tsne <- tsne$Y %>% as_tibble %>% 
  cbind(df_samp %>% select(pheno_type, mem_type, class_type)) %>%
  mutate(mem_type=case_when(is.na(mem_type) ~ 'None', TRUE ~ mem_type))

get_density <- function(x, y){
  d <- densCols(x, y, colramp = colorRampPalette(rev(rainbow(10, end = 4/6))))
  col2rgb(d)[1,] + 1L
}

plot_tsne <- function(df){
  df_ctr <- df %>% group_by(pheno_type) %>% 
    summarize(cx=median(V1), cy=median(V2)) %>% ungroup
  df_pt <- df %>% group_by(pheno_type) %>% 
    mutate(dens=get_density(V1, V2), mem_type=mem_type[1]) %>% ungroup 
  
  ggplot(NULL) + 
    geom_point(data=df_pt, aes(x=V1, y=V2, color=pheno_type, alpha=dens, shape=mem_type)) +
    geom_label(data=df_ctr, aes(x=cx, y=cy, color=pheno_type, label=pheno_type), size=3, alpha=.7) +
    flow_theme
}

df_tsne %>% plot_tsne +
  # Add axis/legend/figure titles
  ggtitle('T Cell TSNE Decomposition') +
  guides(
    colour = guide_legend(title='Primary Phenotype'), 
    shape = guide_legend(title='Memory State'), 
    alpha = 'none'
  ) + xlab('') + ylab('')
```

