---
output: github_document
---

# OMIP 030

```{r global-options, include=FALSE}
library(knitr)
opts_chunk$set(results='hide', warning=FALSE, message=FALSE)
```

```{r libraries}
library(magrittr)
library(tidyverse)
library(fs)
library(janitor)
library(FlowRepositoryR)
library(flowCore)
library(openCyto)
library(ggcyto)
library(gplots)
library(Rtsne)

map <- purrr::map
add <- flowCore::add
filter <- dplyr::filter
```

```{r theme, include=FALSE}
# Set common ggplot theme
flow_theme <- theme_bw() + theme(
  panel.grid.minor.x=element_blank(),
  panel.grid.minor.y=element_blank(),
  panel.grid.major.x=element_blank(),
  panel.grid.major.y=element_blank()
)
```

Load a flow workspace, which will contain all gaiting and cell expression information:

```{r load_workspace}
ws <- openWorkspace('~/analysis/data/omip/omip-030/clean/OMIP_PBMC_clean.wsp')
gs <- parseWorkspace(ws, name='TCells', 
                     path='/home/eczech/analysis/data/omip/omip-030/clean', isNcdf=FALSE)
# Extract single GatingHierarchy from GatingSet
gh <- gs[[1]]
```

```{r summarize_flow_frame, results='markup'}
print(getData(gh))
```

Plot the gating workflow starting from where T Cells are first identified:

```{r show_workflow, fig.width=12}
plot(gs, 'T Cells', width=6, height=1, fontsize=18, shape='rectangle')
```


Identify terminal nodes in the workflow which were, by convention, named with either ```CD4+``` or ```CD8+``` prefixed to their population names:

```{r node_groups}
all_nodes <- getNodes(gs, path=1)
cell_nodes <- getNodes(gs, path=1) %>% str_match(., 'CD[4|8]\\+ .*') %>% discard(is.na)
cell_type_to_class <- function(x) str_detect(x, 'CD4+') %>% ifelse('CD4+', 'CD8+') 
cell_class <- cell_type_to_class(cell_nodes) %>% set_names(cell_nodes)
cell_nodes_th2 <- c('CD4+ Th2g1', 'CD4+ Th2g2')
cell_nodes
```

Show the gating for terminal nodes, which correspond to specific cell types:

```{r leaf_gates, fig.width=12, fig.height=12}
autoplot(gs[[1]], cell_nodes, bins=100, strip.text='gate', axis_inverse_trans=FALSE) + 
  ggcyto_par_set(facet=facet_wrap(~name, scales='free'), limits = list(x=c(0, 225), y=c(0, 225)))
```

Plot the distribution of different cell types noting that many cells are assigned to more than one cell type, and here any duplicate assignments are resolved by counting *all* cells assigned to any one type:

```{r cell_type_distribution, fig.width=12}
cell_count <- getTotal(gh, 'T Cells')

cell_pop <- getPopStats(gs) %>% 
  filter(Population %in% cell_nodes) %>% 
  select(Population, Count) %>% 
  arrange(Count)

cell_pop %>%
  mutate(Percent=Count/cell_count, Class=cell_type_to_class(Population)) %>% 
  arrange(desc(Percent)) %>% 
  mutate(Population=fct_reorder(Population, Percent)) %>%
  ggplot(aes(x=Population, y=Percent, fill=Class)) +
  geom_bar(stat='identity') +
  scale_fill_brewer(palette='Set1') +
  scale_y_continuous(labels = scales::percent) +
  ggtitle('T Cell Type Distribution') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```

To move forward though, determining a single cell type for each cell is done by using the label with
the lowest frequency to maximize the size of populations relating to more rare cells.  Before doing that, here is the frequency of all cell type assignments where multiple assignments are comma separated:

```{r multiple_assignment_distribution, results='asis'}
cell_nodes %>% map(~getIndices(gh, .)) %>% bind_cols %>% 
  apply(., 1, function(x) paste(cell_nodes[x], collapse=', ')) %>% 
  tabyl(var1='assignment') %>%
  arrange(n) %>% adorn_pct_formatting() %>%
  knitr::kable()
```

Choose one cell type for each cell:

```{r extract}
# For now, favor less frequent cell types when assigning single type
get_cell_type <- function(x){
  ctyps <- cell_nodes[x]
  if (is_empty(ctyps)) NA
  # Select rarest cell type if there are multiple
  else cell_pop$Population[which.max(cell_pop$Population %in% ctyps)]
}
cell_types <- cell_nodes %>% map(~getIndices(gh, .)) %>% bind_cols %>% 
  apply(., 1, get_cell_type) 
cell_types <- case_when(cell_types %in% cell_nodes_th2 ~ 'CD4+ Th2', TRUE ~ cell_types)

# Create data frame with measurments and cell type
fr <- getData(gh)
name_map <- parameters(fr)@data %>% filter(str_detect(name, 'Comp'))
name_map <- set_names(name_map$name, name_map$desc)
df <- fr %>% exprs %>% as_tibble %>% 
  select(starts_with('Comp')) %>% 
  rename(!!name_map) %>%
  mutate(type=cell_types) %>%
  filter(!is.na(type)) %>%
  mutate(class=cell_type_to_class(type))
```


Calculate the Kolmogrov Smirnov statistic for each cell type and marker, where the reference sample for any one combination is all *other* T Cells (for the same marker):

```{r ks, fig.width=12, fig.height=8}
dfg <- df %>% group_by(type) %>% do({
  d <- .
  #dp <- df %>% filter(type != d$type[1] & class == d$class[1]) %>% select_if(is.numeric)
  dp <- df %>% filter(type != d$type[1]) %>% select_if(is.numeric)
  dt <- d %>% select_if(is.numeric)
  assertthat::are_equal(colnames(dp), colnames(dt))
  stats <- lapply(colnames(dp), function(col) {
    x <- pull(dt, col)
    y <- pull(dp, col)
    stat <- ks.test(x, y)$statistic[[1]]
    sign(median(x) - median(y)) * stat
  })
  names(stats) <- colnames(dp)
  data.frame(stats)
}) %>% ungroup

dfm <- dfg %>% select(-type) %>% as.matrix
row.names(dfm) <- dfg$type
heatmap.2(
  t(dfm), Rowv=TRUE, Colv=TRUE, 
  col = colorRampPalette(RColorBrewer::brewer.pal(11, 'RdBu'))(16), cexCol=.8, cexRow=.8, 
  margins = c(8, 8), 
  dendrogram = 'col', trace='none', scale='none',
  colsep=1:nrow(dfm), rowsep=1:ncol(dfm),
  density.info='none', sepwidth=c(.05,.05), key.title='KS', key.xlab='', 
  keysize=.5, key.par =list(cex=.3)
)
```

Plot TSNE decomposition:

```{r tsne, fig.width=12, fig.height=8}
set.seed(1)
df_samp <- df %>% sample_n(10000)
tsne <- df_samp %>% select(-type, -class) %>% as.matrix %>% Rtsne
df_tsne <- tsne$Y %>% as_tibble %>%
  mutate(type=df_samp$type, class=df_samp$class)

df_tsne %>% mutate(type=factor(type, levels=sample(unique(type)))) %>%
  ggplot(aes(x=V1, y=V2, color=type, shape=class)) + geom_point() +
  flow_theme
```
